---
title: "methylation"
output: pdf_document
always_allow_html: true
---

```{r setup, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
```

## Introduction

Install the following packages:

```{r packages, echo=FALSE}
library(rmarkdown) # paged table
library(GEOquery) #geo query access
library(dplyr)
library(devtools)
library(ggplot2)
library(limma)
library(Glimma)
library(edgeR)
library(kableExtra)
library(ggrepel)
Sys.setenv(VROOM_CONNECTION_SIZE = 25600000)

```

## Importing the data

```{r importing, message=FALSE, echo=FALSE}

my_id <- "GSE32866"
gse <- getGEO(my_id)
```

Some datasets on GEO may be derived from different microarray platforms. Therefore the object gse is a list of different datasets. You can find out how many were used by checking the length of the gse object. Usually there will only be one platform and the dataset we want to analyse will be the first object in the list (gse[[1]]).

```{r length, message=FALSE}
length(gse)
```

## Extract the data

```{r length2, message=FALSE}
gse <- gse[[1]]
gse
```

## Exploratory analysis

The `exprs` function can retrieve the expression values as a data frame; with one column per-sample and one row per-gene.

```{r, message=FALSE}
pdata= pData(gse) #sample information
edata= exprs(gse) #expression data
fdata = fData(gse) #gene annotation
```

## Inspect the clinical variables

Data submitted to GEO contain sample labels assigned by the experimenters, and some information about the processing protocol. All these data can be extracted by the pData function.











